---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Environmental Health and Justice in 10 California Counties"
subtitle: "https://github.com/tih7/ENV872-DA_FinalProject_Husseini-T.git"
author: "Tamera Husseini"
fontsize: 12pt
mainfont: Times New Roman


---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

getwd()

# Load your packages

library(tidyverse)
library(lubridate)
library(nlme)
library(lme4)
library(corrplot)
library(sf)
library(nationalparkcolors)
library(cowplot)
library(gridExtra)
library(corrplot)
library(here)
library(viridis)
# Set your ggplot theme

my_theme <- theme_grey(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
  
theme_set(my_theme)

# Load your datasets

caliEJscreen <- read.csv("./data/data_raw/cali_ej_screen.csv")

```


# Rationale and Research Questions

The US EPA defines Environmental Justice as *"the fair treatment and meaningful involvement of all people regardless of race, colour, national origin, or income, with respect to the development , implementation and  enforcement of environmental laws, regularions and policies."* Sparked by the protests against the construction of a hazardous waste landfill in Warren County, North Carolina in 1982, the Environmental Justice has slowly taken root in U.S. policy-making. The events in Warren County began to shed light on the disproportionate effect of environmental impacts on low-income communities, communities of colour and indigenous people.  Following the directives of Executive Order 12898 , federal and state authorities have developed online open-source mapping tools to help identify and prioritizing communities experiencing disproportionate environmental burdens. 
 
In the State of California, CalEPA's Office of Environmental Health Hazard Assessment (OEHHA) released the CalEnviroScreen mapping tool in 2013. The CalEnviroScreen data set is based on 20 indicator data categorized broadly into four groups: exposures, environmental effects, sensitive populations and socioeconomic status. CalEnviroScreen's most important feature is its ability to produce rank scores of cumulative impacts for every census tract inthe state (Lee, 2020). More information about CalEnviroScreen can be accessed at https://oehha.ca.gov/calenviroscreen. 

Using the CalEJScreen data set, this study aims to characterise Environmental Justice issues in a sample of 10 Californian counties (Fresno, LA, Madera, Modoc, Monterey, San Bernardina, Sand Diego, San Francisco, Sonoma, Yuba) by exploring the following questions:

1. Which Exposure indicators predict the presence of Sensitive Populations?
2. Which Socioeconomic indicators predict the presence of Sensitive Populations?


\newpage

# Dataset Information

The data set used in this study is an aggregate of environmental, health and socioeconomic indicator data for california 2010 census tract, downloaded from CalEnviroScreen 3.0 (https://oehha.ca.gov/calenviroscreen/maps-data/download-data.) an environmental justice mapping tool. The CalEnviroScreen data set combines 20 indicator data sets categorized broadly as : (1) exposure indicators, (2) enviornmental effects indicators, (3) sensitive population indicators and (4) socio-economic indicators. In addition to the raw indicator data includes a derived Pollution Burden score and Population Characteristic score, which were combined to produce a rank score of cumulative impacts.

The full dataset was filtered down to the indicator variables shown in Table 1. 

```{r echo = FALSE}

var.tbl <- read.csv("./docs/VariablesTable.csv")
knitr::kable(var.tbl, "markdown", caption = "Description of Indicator Variables") 

```


\newpage

# Exploratory Analysis 

```{r echo=FALSE}
#datawrangling
colnames(caliEJscreen)

filtered.data <- caliEJscreen %>% 
  select(Census.Tract,California.County,CES.3.0.Score,Ozone,PM2.5,Diesel.PM,
         Drinking.Water,Pesticides,Tox..Release,Traffic, Pollution.Burden.Score,
         Asthma,Low.Birth.Weight,Cardiovascular.Disease,Education,
         Linguistic.Isolation,Poverty,Unemployment,Housing.Burden,Pop..Char..Score)
head(filtered.data)
colnames(filtered.data)

# selecting counties
summary(filtered.data$California.County)
ten.counties <- filter(filtered.data, California.County %in% c("Los Angeles","San Diego","San Francisco","Sonoma ","Monterey ","Fresno ", "Madera ", "Modoc ", "Yuba ", "San Bernardino"))

summary(ten.counties$California.County)
class(cities$California.County)


```


```{r echo=FALSE, message = FALSE, fig.cap = "California 2010 census tracts"}

# geospatial mapping
cali.map <- st_read("./data/data_spatial/CES3June2018Update.shp")
st_crs(cali.map)
head(cali.map)
colnames(cali.map)

#asthma
ggplot(data = cali.map) +
  geom_sf(aes(fill = asthma)) +
  #geom_sf_text(aes(label = NAME)) +
  theme_bw()

#cardio
ggplot(data = cali.map) +
  geom_sf(aes(fill = cvd)) +
  theme_bw()

#lbw
ggplot(data = cali.map) +
  geom_sf(aes(fill = lbw)) +
  theme_bw()

```



```{r echo=FALSE}
#which counties have the highest rates of asthma, cvd and lbw? 
county.sensitive <- 
  ten.counties %>% 
  group_by(California.County) %>% 
  summarise(mean.asthma = mean(Asthma),
            mean.cvd = mean(Cardiovascular.Disease),
            mean.lbw = mean(Low.Birth.Weight))

summary(county.sensitive)

```

Sensitive populations were explored at the county level. 

```{r fig.cap="Asthma Rates in 10 California Counties"}
# asthma rastes by county
county.box <- ggplot(ten.counties) + 
  geom_boxplot(aes(x= California.County, y=Asthma, colour = California.County)) +
  labs( y = "Asthma (Hospitalization Rate", x="County") +
  theme(axis.text.x  = element_text(angle = 30)) +
  scale_color_viridis(discrete = TRUE) +
  theme(legend.position = " ") +
  ylim(0,200)
county.box

```

```{r fig.cap="Hospitalization rate for Cardiovascular Disease in 10 California counties"}
# cvd rates by county
county.box2 <- ggplot(ten.counties) + 
  geom_boxplot(aes(x= California.County, y=Cardiovascular.Disease, colour = California.County)) +
  labs( y = "Cardiovascular Disease (Hospitalization rate)", x="County") +
  scale_color_viridis(discrete = TRUE) +
  theme(axis.text.x = element_text(angle = 30)) +
  theme(legend.position = " ")
county.box2
```

```{r fig.cap= "Percentage low birth weight in 10 California Counties"}
#lbw by county
county.box3 <- ggplot(ten.counties) + 
  geom_boxplot(aes(x= California.County, y=Low.Birth.Weight, colour = California.County)) +
  labs( y = "Low Birth Weight (%)", x="County") +
  theme(axis.text.x = element_text(angle = 30)) +
  scale_color_viridis(discrete = TRUE) +
  theme(legend.position = " ")
county.box3

```

Because it is typical for socio-economic factors to be related with one another, a Pearson's correlation test was used to test for correlation betweem the socioeconomic factors.  Poverty was strongly positively correlated with with Education (R = 0.82, p<0.01) and Housing Burden (R=0.72, p<0.01) and moderately corelated with Unemployment (R = 0.60, p<0.01) and Linguistic Isolation (R = 0.62, p<0.01). 

```{r echo=FALSE, fig.cap="Correlation matric of socio-economic variables"}

# Correlation between Socio-Economic Indicators
cor.test(caliEJscreen$Poverty, caliEJscreen$Education) #strong correlation 
cor.test(caliEJscreen$Poverty, caliEJscreen$Unemployment) 
cor.test(caliEJscreen$Poverty, caliEJscreen$Housing.Burden) #strong correlation 
cor.test(caliEJscreen$Poverty, caliEJscreen$Linguistic.Isolation)

socioecon.indic <- 
  filtered.data %>% 
  select(Education:Housing.Burden) %>% 
  na.omit()

socioecon.cor <-cor(socioecon.indic)
#corrplot(socioecon.cor, method = "ellipse")  
corrplot.mixed(socioecon.cor, upper = "ellipse", tl.pos = "d")

```
\newpage

# Analysis
## Question 1: Which Exposure indicators predict the presence of Sensitive Population?

A linear regression model was performed to test which Pollution Burden indicators predict the presence of Sensitive Populations. Three different response variables - Asthma  rates, Cardiovascular disease rates and low birth weight - were tested against a suite of Pollution Burden explanatory variables. 


```{r echo = FALSE}

# Asthma ~ Exposure

colnames(ten.counties)

asthma.mdl <- lm(Asthma ~ Ozone + PM2.5 + Diesel.PM + Traffic + Drinking.Water + Tox..Release + Pesticides, data = ten.counties) 
summary(asthma.mdl)

asthma.mdl2 <-lm (Asthma ~ Ozone + Diesel.PM + Drinking.Water + Tox..Release, data = ten.counties)
summary(asthma.mdl2)
plot(asthma.mdl2)

asthma.mdl3 <-lm (Asthma ~ Ozone + Diesel.PM + Drinking.Water, data = ten.counties)
summary(asthma.mdl3)

# Plots
#2 Asthma ~ ozone

asthma.o3 <- ggplot(ten.counties, aes(x=Ozone, y= Asthma, colour= California.County)) +
  geom_point() +
  scale_color_viridis(discrete = TRUE)+
  geom_smooth(method = "lm", colour= "black", se=FALSE)
asthma.o3

#3 Asthma ~ Diesel

asthma.dis <- ggplot(ten.counties, aes(x=Diesel.PM, y= Asthma, colour= California.County)) +
  geom_point() +
  scale_color_viridis(discrete = TRUE)+
  geom_smooth(method = "lm", colour= "black", se=FALSE)
asthma.dis

#4 Asthma ~ drinking water

asthma.dw <- ggplot(ten.counties, aes(x=Drinking.Water, y= Asthma, colour= California.County)) +
  geom_point() +
  scale_color_viridis(discrete = TRUE)+
  geom_smooth(method = "lm", colour= "black", se=FALSE)
asthma.dw

```

```{r}
# Cardiovascular disease ~ Exposure

cardio.mdl <- lm(Cardiovascular.Disease ~ Ozone + PM2.5 + Diesel.PM + Traffic + Drinking.Water + Tox..Release + Pesticides, data = filtered.data)
summary(cardio.mdl)

cardio.mdl2 <- lm(Cardiovascular.Disease ~ Ozone + Diesel.PM + Drinking.Water + Traffic, data= filtered.data)
summary(cardio.mdl2)
plot(cardio.mdl2)

# Birth weight ~ Exposure

bw.mdl1 <- lm(Low.Birth.Weight ~ Ozone + PM2.5 + Diesel.PM + Traffic + Drinking.Water + Tox..Release + Pesticides, data = caliEJscreen) 
summary (bw.mdl1)

bw.md2 <- lm(Low.Birth.Weight ~ Ozone + PM2.5 + Diesel.PM + Traffic  , data = filtered.data)
summary(bw.md2)

bw.mdl3 <- lm(Low.Birth.Weight ~ Ozone + PM2.5 + Diesel.PM  , data = filtered.data)
summary(bw.mdl3)

```

```{r echo = FALSE}

# asthma ~ poverty, CA
asthma.poverty.ca <- ggplot(filtered.data,aes(y=Asthma, x= Poverty)) +
  geom_point(alpha = 0.7) +
  ylab("Cardiovascular Disease (Rate)")+
  geom_smooth(method="lm", se=FALSE, colour = "black")
asthma.poverty.ca
```

```{r, echo=FALSE, warning=FALSE}

# cardio ~ poverty, CA
cardio.poverty.ca <- ggplot(filtered.data,aes(y=Cardiovascular.Disease, x= Poverty)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method="lm", se=FALSE, col = "black")
cardio.poverty.ca
```

```{r echo=FALSE}
visualisation <- grid.arrange(asthma.poverty.ca, cardio.poverty.ca, ncol=1)
visualisation

#pm2.5 ~ poverty 
pm25.poverty.ca <- ggplot(caliEJscreen,aes(y=PM2.5, x= Poverty)) +
  geom_point(alpha = 0.7) +
  ylab("PM 2.5")+
  geom_smooth(method="lm", se=FALSE, colour = "black")
pm25.poverty.ca

```


## Question 2: Which Socioeconomic indicators predict the presence of Sensitive Populations?

Because Poverty was found to be positively corelated with other socio-economic factors in the dataset, it was used as a blanket factor to test the relationship between socio-economic indicators and sensitive population indicators. Poverty was found to be significantly related to Asthma (R^2^ = 0.238, d.f. =), cardiovascular disease (), and low Birth Weight.  

```{r echo=FALSE, fig.cap="The relationship between poverty and sensitive population indicators"}

#LR models

# Asthma ~ Poverty
asthma.pv <- lm(Asthma ~ Poverty, data = filtered.data )
summary(asthma.pv)

# Cardio ~ Poverty
cardio.pv <- lm(Cardiovascular.Disease ~ Poverty, data = filtered.data)
summary(cardio.pv)

# BW ~ Poverty 
bw.pv <- lm(Low.Birth.Weight ~ Poverty, data = filtered.data)
summary(bw.pv)

#scatterplot

#1 Asthma ~ Poverty
asthma.pov.plot<- ggplot(ten.counties, aes(y=Asthma, x= Poverty, colour= California.County)) +
  geom_point(alpha = 0.7) +
  ylab("Asthma (ER visits per 10,000)")+
  scale_color_viridis(discrete = TRUE) +
  geom_smooth(method="lm", se=FALSE, colour = "black")+
  theme(legend.position = " ")
#asthma.pov.plot

#2 cvd ~ poverty
cvd.pov.plot<- ggplot(ten.cali.counties, aes(y=Cardiovascular.Disease, x= Poverty, colour= California.County)) +
  geom_point(alpha = 0.7) +
  ylab("Cardiovascular Disease (ER visits per 10,000)")+
  scale_color_viridis(discrete = TRUE) +
  geom_smooth(method="lm", se=FALSE, colour = "black")+
  theme(legend.position = " ")
#cvd.pov.plot

#lbw ~ poverty
lbw.pov.plot<- ggplot(ten.cali.counties, aes(y=Low.Birth.Weight, x= Poverty, colour= California.County)) +
  geom_point(alpha = 0.7) +
  ylab("Low Birth Weight (%)")+
  scale_color_viridis(discrete = TRUE) +
  geom_smooth(method="lm", se=FALSE, colour = "black")+
  theme(legend.position = " ")
#lbw.pov.plot


#legend
legend<- get_legend(asthmaplot + theme(legend.position = "bottom", legend.direction = "vertical"))

viz<- plot_grid(asthma.pov.plot, lbw.pov.plot,cvd.pov.plot, legend,nrow =1)
viz
```


\newpage

# Summary and Conclusions
The connection between, poverty (and other socioeconomic factors) and environmental health is evident in this data set. 

\newpage

# References

Lee, C. (2020) A Game Changer in the Making? Lessions from States advancing Environmental Justice through Mapp and Cumulative Impact Strategies. Environmental Law Institute. 
