---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Environmental Health and Justice in California"
subtitle: "https://github.com/tih7/ENV872-DA_FinalProject_Husseini-T.git"
author: "Tamera Husseini"
fontsize: 12pt
mainfont: Times New Roman


---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

getwd()

# Load your packages

library(tidyverse)
library(lubridate)
library(nlme)
library(lme4)
library(corrplot)
library(sf)
library(nationalparkcolors)
library(cowplot)
library(gridExtra)
# Set your ggplot theme

my_theme <- theme_grey(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
  
theme_set(my_theme)

# Load your datasets

caliEJscreen <- read.csv("./data/data_raw/cali_ej_screen.csv")

```


# Rationale and Research Questions

The US EPA defines Environmental Justice as *"the fair treatment and meaningful involvement of all people regardless of race, colour, national origin, or income, with respect to the development , implementation and  enforcement of environmental laws, regularions and policies."* Sparked by the protests against the construction of a hazardous waste landfill in Warren County, North Carolina in 1982, the Environmental Justice has slowly taken root in U.S. policy-making. The events in Warren County began to shed light on the disproportionate effect of environmental impacts on low-income communities, communities of colour and indigenous people.  Following the directives of Executive Order 12898 , federal and state authorities have developed online open-source mapping tools to help identify and prioritizing communities experiencing disproportionate environmental burdens. 
 
In the State of California, CalEPA's Office of Environmental Health Hazard Assessment (OEHHA) released the CalEnviroScreen mapping tool in 2013. The CalEnviroScreen data set is based on 20 indicator data categorized broadly into four groups: exposures, environmental effects, sensitive populations and socioeconomic status. CalEnviroScreen's most important feature is its ability to produce rank scores of cumulative impacts for every census tract inthe state (Lee, 2020). More information about CalEnviroScreen can be accessed at https://oehha.ca.gov/calenviroscreen. 

Using the CalEJScreen data set, this study aims to characterise Environmental Justice in the State of California by addressing the following questions:

1. Which Exposure and Socio-economic indicators predict the presence of Sensitive Population?
2. Which counties have the highest median rates of Asthma, Cardiovascular Disease and Low Birth Weight (Sensitive Population Indicators)

\newpage

# Dataset Information

The data set used in this study is an aggregate of environmental, health and socioeconomic indicator data for the State of California, downloaded from CalEnviroScreen 3.0, an environmental justice mapping tool developed by the California Office of Environmental Health Hazard Assessment (OEHHA). The CalEnviroScreen data set combines 20 indicator data sets categorized broadly as : (1) exposure indicators, (2) enviornmental effects indicators, (3) sensitive population indicators and (4) socio-economic indicators. In addition to the indicator data, 

derive a rank score. 

An area with a high score is one that experiences a much higher pollution burden than areas with low scores. The tool was designed to help identify California communities that are most affected by many sources of pollution. The data is divided along 2010 census tracts and is up-to-date as of 2018. 

More information about CalEnviroScreen 3.0 can be found here: https://oehha.ca.gov/calenviroscreen/about-calenviroscreen

Geographic and Demographic Data:

**Census Tract**: Census tract ID from the 2010 census.
* Total Population: 2010 census tract population
**California County**: California county name in which the tract is located
* ZIP: Postal zip code in which tract is located
* Nearby City: closest city to census tract to help approximate census tract location
* Longtitude:Longtitude at the center of the census tract in decimal degrees
* Latitude: Latitude of the census in decimal degrees

CaliEnvironScreen Rank Score

* CES 3.0 Score: CalEnviroScreen Score, Pollution Score multiplied by Population Characteristics Score
* CES 3.0 Percentile: Percentile of the CalEnviroScreen score
* CES 3.0 Percentile Range: Percentile of the CalEnviroScreen score
* SB 535 Disadvantaged Community: Disadvantaged communities list from:
https://oehha.ca.gov/calenviroscreen/sb535
http://www.calepa.ca.gov/EnvJustice/GHGInvest/ 


Exposure Indicators: 

* Ozone: Amount of daily max. 8hr Ozone concentration
* PM2.5: Annual mean PM 2.5 concentration
* Diesel PM: Diesel PM emission from on-road and non-road sources
* Drinking Water: Drinking water contaminant index for selected contaminnats
* Pesticides: Total pounds of selected active pesticide ingredients (filtered for hazard and volatility) used in production-agriculture per square mile in the census tract.
* Tox. Release: Toxicity-weighted concentrations of modeled chemical releases to air from facility emissions and off-site incineration (from RSEI)
* Traffic: Traffic density, in vehicle-kilometers per hour per road length, within 150 meters of the census tract boundary

Environmental Effects Indicators:

* Cleanup Sites: Traffic density, in vehicle-kilometers per hour per road length, within 150 meters of the census tract boundary
* Groundwater Threats:Groundwater threats, sum of weighted GeoTracker leaking underground storage tank sites  within buffered distances to populated blocks of census tracts
* Haz. Waste: Sum of weighted hazardous waste facilities and large quantity generators within buffered distances to populated blocks of census tracts.
*Imp. Water Bodies: Impaired water bodies, sum of number of pollutants across all impaired water bodies within buffered distances to populated blocks of census tracts. 
* Solid Waste:Sum of weighted hazardous waste facilities and large quantity generators within buffered distances to populated blocks of census tracts

Sensitive Population Indicators:

* Asthma: Age-adjusted rate of emergency department visits for asthma
* Low Birth Weight: Percent low birth weigth
* Cardiovascular Disease: Age-adjusted rate of emergency dept. visits for heart attacks per 10,000.

Socio-economic Factor Indicators:

* Education: Percent of population living below two times the federal poverty level
* Linguistic Isolation: Percent limited English speaking households.
* Poverty: Percent of population living below two times the federal poverty level.
* Unemployment: Percent of the population over the age of 16 that is unemployed and eligible for the labor force.
* Housing Burden: Percent housing burdened low income households. 



\newpage

# Exploratory Analysis 
```{r, fig.height = 6, fig.width = 4, warning = FALSE, fig.cap= "The effects of poverty on rates of asthma and cardiovascular disease in California census tract populations." }


#USE THIS ONE FOR VISUALISATION
# asthma ~ poverty, CA
asthma.poverty.ca <- ggplot(caliEJscreen, aes(y=Asthma, x= Poverty)) +
  geom_point(alpha = 0.7, colour = everglades[2]) +
  ylab("Asthma (Rate)")+
  xlab("") +
  geom_smooth(method="lm", se=FALSE, colour = "black")


# cardio ~ poverty, CA
cardio.poverty.ca <- ggplot(caliEJscreen,aes(y=Cardiovascular.Disease, x= Poverty)) +
  geom_point(alpha = 0.7, col = everglades[3]) +
  ylab("Cardiovascular Disease (Rate)")+
  xlab("")+
  geom_smooth(method="lm", se=FALSE, col = "black")

# bw ~ poverty, CA
bw.poverty.ca <- ggplot(caliEJscreen,aes(y=Low.Birth.Weight, x= Poverty)) +
  geom_point(alpha = 0.7, col = everglades[4]) + 
  ylab("Low Birth Weight")+
  geom_smooth(method="lm", se=FALSE, col = "black")

visualisation <- grid.arrange(asthma.poverty.ca, cardio.poverty.ca,bw.poverty.ca, ncol=1)

```



\newpage

# Analysis

```{r}
# Mapping


```



## Question 1: Which Exposure indicators predict the presence of Sensitive Population?


A linear regression model was performed to test which Pollution Burden indicators predict the presence of Sensitive Populations. Three different response variables - Asthma  rates, Cardiovascular disease rates and low birth weight - were tested against a suite of Pollution Burden explanatory variables. 


```{r}
# Correlation between Socio-Economic Indicators
cor.test(caliEJscreen$Poverty, caliEJscreen$Education) #strong correlation 
cor.test(caliEJscreen$Poverty, caliEJscreen$Unemployment) 
cor.test(caliEJscreen$Poverty, caliEJscreen$Housing.Burden) #strong correlation 
cor.test(caliEJscreen$Poverty, caliEJscreen$Linguistic.Isolation)
cor.test(caliEJscreen$Poverty, caliEJscreen$Education) #strong correlation

# Asthma

colnames(caliEJscreen)

asthma.mdl <- lm(Asthma ~ Ozone + PM2.5 + Diesel.PM + Traffic + Drinking.Water + Tox..Release + Pesticides, data =caliEJscreen) 
summary(asthma.mdl)

asthma.mdl2 <-lm (Asthma ~ Ozone + Diesel.PM + Traffic + Drinking.Water + Tox..Release, data =caliEJscreen)
summary(asthma.mdl2)
plot(asthma.mdl2)

# Cardiovascular disease 

cardio.mdl <- lm(Cardiovascular.Disease ~ Ozone + PM2.5 + Diesel.PM + Traffic + Drinking.Water + Tox..Release + Pesticides, data = caliEJscreen)
summary(cardio.mdl)

cardio.mdl2 <- lm(Cardiovascular.Disease ~ Ozone + Diesel.PM + Drinking.Water + Traffic, data=caliEJscreen)
summary(cardio.mdl2)
plot(cardio.mdl2)

# Birth weight

bw.mdl1 <- lm(Low.Birth.Weight ~ Ozone + PM2.5 + Diesel.PM + Traffic + Drinking.Water + Tox..Release + Pesticides, data = caliEJscreen) 
summary (bw.mdl1)

bw.md2 <- lm(Low.Birth.Weight ~ Ozone + PM2.5 + Diesel.PM + Traffic  , data = caliEJscreen)
summary(bw.md2)

bw.mdl3 <- lm(Low.Birth.Weight ~ Ozone + PM2.5 + Diesel.PM  , data = caliEJscreen)
summary(bw.mdl3)



```

## Question 2: Which socio-

or

## Question 2: Which counties have the highest median rate of Asthma, Cardiovascular and Low Birth Weight? 

\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
