---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Environmental Health and Justice in California"
subtitle: "https://github.com/tih7/ENV872-DA_FinalProject_Husseini-T.git"
author: "Tamera Husseini"
fontsize: 12pt
mainfont: Times New Roman


---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

getwd()

# Load your packages

library(tidyverse)
library(lubridate)
library(nlme)
library(lme4)
library(corrplot)
library(sf)
library(nationalparkcolors)
library(cowplot)
library(gridExtra)
# Set your ggplot theme

my_theme <- theme_grey(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
  
theme_set(my_theme)

# Load your datasets

caliEJscreen <- read.csv("./data/data_raw/cali_ej_screen.csv")

```


# Rationale and Research Questions

The US EPA defines Environmental Justice as *"the fair treatment and meaningful involvement of all people regardless of race, colour, national origin, or income, with respect to the development , implementation and  enforcement of environmental laws, regularions and policies."* Sparked by the protests against the construction of a hazardous waste landfill in Warren County, North Carolina in 1982, the Environmental Justice has slowly taken root in U.S. policy-making. The events in Warren County began to shed light on the disproportionate effect of environmental impacts on low-income communities, communities of colour and indigenous people.  Following the directives of Executive Order 12898 , federal and state authorities have developed online open-source mapping tools to help identify and prioritizing communities experiencing disproportionate environmental burdens. 
 
In the State of California, CalEPA's Office of Environmental Health Hazard Assessment (OEHHA) released the CalEnviroScreen mapping tool in 2013. The CalEnviroScreen data set is based on 20 indicator data categorized broadly into four groups: exposures, environmental effects, sensitive populations and socioeconomic status. CalEnviroScreen's most important feature is its ability to produce rank scores of cumulative impacts for every census tract inthe state (Lee, 2020). More information about CalEnviroScreen can be accessed at https://oehha.ca.gov/calenviroscreen. 

Using the CalEJScreen data set, this study aims to characterise Environmental Justice in the State of California by addressing the following questions:

1. Which Exposure and Socio-economic indicators predict the presence of Sensitive Population?
2. Which counties have the highest median rates of Asthma, Cardiovascular Disease and Low Birth Weight (Sensitive Population Indicators)

\newpage

# Dataset Information

The data set used in this study is an aggregate of environmental, health and socioeconomic indicator data for california 2010 census tract, downloaded from CalEnviroScreen 3.0 (https://oehha.ca.gov/calenviroscreen/maps-data/download-data.) an environmental justice mapping tool. The CalEnviroScreen data set combines 20 indicator data sets categorized broadly as : (1) exposure indicators, (2) enviornmental effects indicators, (3) sensitive population indicators and (4) socio-economic indicators. In addition to the raw indicator data includes a derived Pollution Burden score and Population Characteristic score, which were combined to produce a rank score of cumulative impacts.

The full dataset was filtered down to the indicator variables shown in Table 1.

```{r, echo = FALSE}

var.tbl <- read.csv("./docs/VariablesTable.csv")
knitr::kable(var.tbl, "markdown", caption = "Description of Indicator Variables") 

```


\newpage

# Exploratory Analysis 

Because it is not uncommon for socio-economic are commonly related with one another, 
```{r}
# Correlation between Socio-Economic Indicators
cor.test(caliEJscreen$Poverty, caliEJscreen$Education) #strong correlation 
cor.test(caliEJscreen$Poverty, caliEJscreen$Unemployment) 
cor.test(caliEJscreen$Poverty, caliEJscreen$Housing.Burden) #strong correlation 
cor.test(caliEJscreen$Poverty, caliEJscreen$Linguistic.Isolation)
cor.test(caliEJscreen$Poverty, caliEJscreen$Education) #strong correlation
```


```{r, fig.height = 6, fig.width = 4, warning = FALSE, fig.cap= "The effects of poverty on rates of asthma and cardiovascular disease in California census tract populations." }


#USE THIS ONE FOR VISUALISATION
# asthma ~ poverty, CA
asthma.poverty.ca <- ggplot(caliEJscreen, aes(y=Asthma, x= Poverty)) +
  geom_point(alpha = 0.7) +
  ylab("Asthma (Rate)")+
  xlab("") +
  geom_smooth(method="lm", se=FALSE, colour = "black")


# cardio ~ poverty, CA
cardio.poverty.ca <- ggplot(caliEJscreen,aes(y=Cardiovascular.Disease, x= Poverty)) +
  geom_point(alpha = 0.7) +
  ylab("Cardiovascular Disease (Rate)")+
  xlab("")+
  geom_smooth(method="lm", se=FALSE, col = "black")

# bw ~ poverty, CA
bw.poverty.ca <- ggplot(caliEJscreen,aes(y=Low.Birth.Weight, x= Poverty)) +
  geom_point(alpha = 0.7) + 
  ylab("Low Birth Weight")+
  geom_smooth(method="lm", se=FALSE, col = "black")

visualisation <- grid.arrange(asthma.poverty.ca, cardio.poverty.ca,bw.poverty.ca, ncol=1)

```



\newpage

# Analysis

```{r}
# Mapping

cali.map <- st_read("./data/data_spatial/CES3June2018Update.shp")
st_crs(cali.map)

here()
```



## Question 1: Which Exposure indicators predict the presence of Sensitive Population?


A linear regression model was performed to test which Pollution Burden indicators predict the presence of Sensitive Populations. Three different response variables - Asthma  rates, Cardiovascular disease rates and low birth weight - were tested against a suite of Pollution Burden explanatory variables. 


```{r}


# Asthma

colnames(caliEJscreen)

asthma.mdl <- lm(Asthma ~ Ozone + PM2.5 + Diesel.PM + Traffic + Drinking.Water + Tox..Release + Pesticides, data =caliEJscreen) 
summary(asthma.mdl)

asthma.mdl2 <-lm (Asthma ~ Ozone + Diesel.PM + Traffic + Drinking.Water + Tox..Release, data =caliEJscreen)
summary(asthma.mdl2)
plot(asthma.mdl2)

# Cardiovascular disease 

cardio.mdl <- lm(Cardiovascular.Disease ~ Ozone + PM2.5 + Diesel.PM + Traffic + Drinking.Water + Tox..Release + Pesticides, data = caliEJscreen)
summary(cardio.mdl)

cardio.mdl2 <- lm(Cardiovascular.Disease ~ Ozone + Diesel.PM + Drinking.Water + Traffic, data=caliEJscreen)
summary(cardio.mdl2)
plot(cardio.mdl2)

# Birth weight

bw.mdl1 <- lm(Low.Birth.Weight ~ Ozone + PM2.5 + Diesel.PM + Traffic + Drinking.Water + Tox..Release + Pesticides, data = caliEJscreen) 
summary (bw.mdl1)

bw.md2 <- lm(Low.Birth.Weight ~ Ozone + PM2.5 + Diesel.PM + Traffic  , data = caliEJscreen)
summary(bw.md2)

bw.mdl3 <- lm(Low.Birth.Weight ~ Ozone + PM2.5 + Diesel.PM  , data = caliEJscreen)
summary(bw.mdl3)



```

## Question 2: Which socio-

or

## Question 2: Which counties have the highest median rate of Asthma, Cardiovascular and Low Birth Weight? 

\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
